---
title: "STAT452 Final Project: Wine Quality and Experimental Design"
author:
  - "Tr\u1EA7n L\u00EA Anh Tu\u1EA5n (24125107)"
  - "Nguy\u1EC5n B\u1EA3o Minh Tri\u1EBFt (24125047)"
  - "Nguy\u1EC5n \u0110\u00ECnh Thi\u00EAn L\u1ED9c (24125093)"
  - "Nguy\u1EC5n H\u1ED3ng T\u1EA5n T\u00E0i (24125078)"
  - "L\u00EA Minh Thu\u1EADn (24125105)"
date: "Academic Year 2025--2026"
bibliography:
  - references.bib
  - ../finals/part1/report/references.bib
link-citations: true
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
    toc: true
    toc_depth: 2
    fig_caption: true
    keep_tex: false
    includes:
      in_header: ../finals/part1/report/header.tex
fontsize: 10pt
geometry: margin=0.78in
---

\begin{abstract}
This report currently completes Part 1 of the Group 8 final project. We predict
red Vinho Verde quality from 11 physicochemical measurements with a locked,
leakage-controlled comparison of OLS, Ridge, Lasso, and Elastic Net. Part 2,
the separate experimental-design extension, remains explicitly pending.
\end{abstract}

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE, results = "asis"
)
source("../R/setup.R")
knitr::opts_knit$set(root.dir = PROJECT_ROOT)

summary_file <- file.path(PROJECT_ROOT, "output", "analysis_summary.RData")
if (!file.exists(summary_file)) {
  stop("Run source('analysis/00_run_all.R') before rendering the report.")
}
load(summary_file)

holdout_table <- analysis_summary$holdout$performance
locked_model <- analysis_summary$model_lock$Locked_model[[1L]]
locked_row <- holdout_table[holdout_table$Model == locked_model, ]
best_row <- holdout_table[which.min(holdout_table$Holdout_RMSE), ]
top_correlations <- analysis_summary$correlations[seq_len(4L), ]
max_vif_row <- analysis_summary$feature_screening[
  which.max(analysis_summary$feature_screening$VIF),
]
selected_lasso <- subset(analysis_summary$lasso_selection, Selected == "Yes")

fmt <- function(x, digits = 3L) formatC(x, digits = digits, format = "f")
pretty_feature <- function(x) gsub("_", " ", x)
```

# Part 1: Wine Quality (Red)

```{r part1-sections}
section_files <- file.path(
  PROJECT_ROOT, "finals", "part1", "report", "sections",
  c(
    "01_exploratory_data_analysis.Rmd",
    "02_data_cleaning.Rmd",
    "03_feature_selection.Rmd",
    "04_regularized_models.Rmd",
    "05_evaluation_conclusion.Rmd"
  )
)
for (section in section_files) {
  cat(
    knitr::knit_child(section, envir = knitr::knit_global(), quiet = TRUE),
    sep = "\n"
  )
}
```

\clearpage

# Part 2: Experimental-design extension

TODO: Select and justify the separate dataset; state hypotheses; complete the
two-factor ANOVA or \(2^k\) factorial analysis; check assumptions; report
interactions and post-hoc comparisons where justified; and interpret practical
significance and limitations.

# Contributions

TODO: Replace the workstream ownership plan with a statement of the work each
member actually completed and reviewed.

```{r appendix}
appendix_file <- file.path(
  PROJECT_ROOT, "finals", "part1", "report", "sections", "06_appendix.Rmd"
)
cat(
  knitr::knit_child(
    appendix_file, envir = knitr::knit_global(), quiet = TRUE
  ),
  sep = "\n"
)
```

\newpage

# References {-}
