---
title: "STAT452 Final Project: Wine Quality and Experimental Design"
author:
  - "Tr\u1EA7n L\u00EA Anh Tu\u1EA5n (24125107)"
  - "Nguy\u1EC5n B\u1EA3o Minh Tri\u1EBFt (24125047)"
  - "Nguy\u1EC5n \u0110\u00ECnh Thi\u00EAn L\u1ED9c (24125093)"
  - "Nguy\u1EC5n H\u1ED3ng T\u1EA5n T\u00E0i (24125078)"
  - "L\u00EA Minh Thu\u1EADn (24125105)"
date: "Academic Year 2025--2026"
bibliography:
  - references.bib
  - ../finals/part1/report/references.bib
  - ../finals/part2/report/references.bib
link-citations: true
output:
  pdf_document:
    latex_engine: xelatex
    number_sections: true
    toc: true
    toc_depth: 2
    fig_caption: true
    keep_tex: true
    includes:
      in_header: ../finals/part1/report/header.tex
fontsize: 10pt
geometry: margin=0.78in
---

\begin{abstract}
This report completes both components of the Group 8 final project. Part 1
predicts red Vinho Verde quality from 11 physicochemical measurements with a
locked, leakage-controlled comparison of OLS, Ridge, Lasso, and Elastic Net.
Part 2 uses an observational two-factor analysis of student mathematics scores
by test preparation and lunch category, with diagnostics, robust sensitivity
checks, and explicitly non-causal interpretation.
\end{abstract}

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE, results = "asis"
)
source("../R/setup.R")
knitr::opts_knit$set(root.dir = PROJECT_ROOT)

summary_file <- file.path(PROJECT_ROOT, "output", "analysis_summary.RData")
if (!file.exists(summary_file)) {
  stop("Run source('analysis/00_run_all.R') before rendering the report.")
}
load(summary_file)

holdout_table <- analysis_summary$holdout$performance
locked_model <- analysis_summary$model_lock$Locked_model[[1L]]
locked_row <- holdout_table[holdout_table$Model == locked_model, ]
best_row <- holdout_table[which.min(holdout_table$Holdout_RMSE), ]
top_correlations <- analysis_summary$correlations[seq_len(4L), ]
max_vif_row <- analysis_summary$feature_screening[
  which.max(analysis_summary$feature_screening$VIF),
]
selected_lasso <- subset(analysis_summary$lasso_selection, Selected == "Yes")

part2_file <- file.path(PROJECT_ROOT, "output", "part2_summary.RData")
if (!file.exists(part2_file)) {
  stop("Part 2 output is missing. Run source('analysis/00_run_all.R').")
}
load(part2_file)

p2_find <- function(label) {
  part2_summary$key_findings[
    part2_summary$key_findings$Finding == label, , drop = FALSE
  ]
}
p2_preparation <- p2_find("Test-preparation association")
p2_lunch <- p2_find("Lunch-group association")
p2_interaction <- p2_find("Preparation-by-lunch interaction")
p2_shapiro <- subset(part2_summary$assumptions,
                     Assumption == "Residual normality")
p2_bf <- subset(part2_summary$assumptions,
                Assumption == "Homogeneity of variance")
p2_influence <- subset(part2_summary$assumptions,
                       Assumption == "Influential observations")
p2_max_cook <- p2_influence$Statistic[[1L]]
p2_influence_count <- as.integer(sub(
  ".*4/n: ([0-9]+);.*", "\\1", p2_influence$Assessment[[1L]]
))
p1_cook <- subset(analysis_summary$influence_audit,
                  Diagnostic == "Cook's distance")


fmt <- function(x, digits = 3L) formatC(x, digits = digits, format = "f")
pretty_feature <- function(x) gsub("_", " ", x)
p2_fmt <- fmt
p2_interval <- function(lower, upper) {
  sprintf("[%s, %s]", p2_fmt(lower, 2L), p2_fmt(upper, 2L))
}
```

# Introduction and analysis protocol

This project addresses two complementary questions. Part 1 asks how accurately
red-wine sensory quality can be predicted from routine physicochemical
measurements. Part 2 asks whether mathematics scores differ across observed
test-preparation and lunch groups, including whether the two associations
interact. The analyses follow the course brief's emphasis on justified choices,
reproducibility, and honest limitations.

For Part 1, RMSE is primary because large one-wine prediction errors are costly;
MAE expresses typical error directly in quality-score points, and $R^2$ measures
improvement over the holdout mean benchmark. Model selection uses fixed five-fold
cross-validation, while the stratified 20% holdout is opened once [@Hastie2009].

# Part 1: Wine Quality (Red)

```{r part1-sections}
section_files <- file.path(
  PROJECT_ROOT, "finals", "part1", "report", "sections",
  c(
    "01_exploratory_data_analysis.Rmd",
    "02_data_cleaning.Rmd",
    "03_feature_selection.Rmd",
    "04_regularized_models.Rmd",
    "05_evaluation_conclusion.Rmd"
  )
)
for (section in section_files) {
  cat(
    knitr::knit_child(section, envir = knitr::knit_global(), quiet = TRUE),
    sep = "\n"
  )
}
```

\clearpage

# Part 2: Experimental-design extension

```{r part2-sections}
part2_sections <- file.path(
  PROJECT_ROOT, "finals", "part2", "report", "sections",
  c(
    "01_dataset_design.Rmd",
    "02_descriptive_statistics.Rmd",
    "03_factorial_anova.Rmd",
    "04_diagnostics_posthoc.Rmd",
    "05_conclusion_limitations.Rmd"
  )
)
for (section in part2_sections) {
  cat(
    knitr::knit_child(section, envir = knitr::knit_global(), quiet = TRUE),
    sep = "\n"
  )
}
```

# Contributions and review

The repository history, module ownership, and completed handoffs record the
following contributions:

- **Nguyễn Đình Thiên Lộc:** project framework and reproducibility design; Part
  1 data preparation/EDA; Part 2 data audit, design justification, and hypotheses.
- **Trần Lê Anh Tuấn:** Part 1 cleaning and preprocessing; Part 2 descriptive
  statistics and visualizations.
- **Lê Minh Thuận:** Part 1 feature selection and baselines; Part 2 effect-coded
  factorial ANOVA and effect sizes.
- **Nguyễn Bảo Minh Triết:** Part 1 regularized modelling and canonical
  integration; Part 2 interpretation and report integration.
- **Nguyễn Hồng Tấn Tài:** Part 1 locked holdout evaluation; Part 2 diagnostics,
  robust sensitivity, and multiplicity-adjusted follow-up comparisons.

The final integration review checked every code-to-table/figure handoff, the
holdout boundary, contrast directions, diagnostics, citations, and PDF layout.
Dataset discovery is not counted as a contribution.

```{r appendix}
appendix_file <- file.path(
  PROJECT_ROOT, "finals", "part1", "report", "sections", "06_appendix.Rmd"
)
cat(
  knitr::knit_child(
    appendix_file, envir = knitr::knit_global(), quiet = TRUE
  ),
  sep = "\n"
)
```

\newpage

# References {-}
