---
title: "STAT452 Final Project — Part 1"
subtitle: "Regularized Regression for Red Wine Quality"
author: |
  **Group 08**  \\
  Trần Lê Anh Tuấn (24125107); Nguyễn Bảo Minh Triết (24125047);  \\
  Nguyễn Đình Thiên Lộc (24125093); Nguyễn Hồng Tấn Tài (24125078);  \\
  Lê Minh Thuận (24125105)
date: "Academic Year 2025–2026"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
    keep_tex: false
    includes:
      in_header: header.tex
geometry: margin=0.78in
fontsize: 10pt
bibliography: references.bib
link-citations: true
params:
  run_pipeline: true
---

```{r setup, include=FALSE}
input_file <- normalizePath(knitr::current_input(dir = TRUE), winslash = "/",
                            mustWork = TRUE)
source(file.path(dirname(input_file), "..", "setup.R"))
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      results = "asis")
knitr::opts_knit$set(root.dir = PART1_ROOT)
if (isTRUE(params$run_pipeline)) {
  source(file.path(PART1_ROOT, "R", "00_run_all.R"),
         local = new.env(parent = globalenv()))
}
load(file.path(OUTPUT_DIR, "analysis_summary.RData"))

holdout_table <- analysis_summary$holdout$performance
locked_model <- analysis_summary$model_lock$Locked_model[[1]]
locked_row <- holdout_table[holdout_table$Model == locked_model, ]
best_row <- holdout_table[which.min(holdout_table$Holdout_RMSE), ]
top_correlations <- analysis_summary$correlations[1:4, ]
max_vif_row <- analysis_summary$feature_screening[
  which.max(analysis_summary$feature_screening$VIF),
]
selected_lasso <- subset(analysis_summary$lasso_selection, Selected == "Yes")

fmt <- function(x, digits = 3) formatC(x, digits = digits, format = "f")
pretty_feature <- function(x) gsub("_", " ", x)
```

\begin{abstract}
We predict the expert quality score of Portuguese red Vinho Verde from 11
physicochemical measurements. The workflow removes exact duplicates before a
fixed stratified split, performs all exploratory and preprocessing decisions on
the training partition, and compares OLS with Ridge, Lasso, and Elastic Net
using shared fold-clean cross-validation. The pre-holdout choice was
**`r locked_model`** (CV RMSE `r fmt(analysis_summary$model_lock$CV_RMSE[[1]])`).
On `r analysis_summary$config$holdout_rows` untouched observations it achieved
RMSE `r fmt(locked_row$Holdout_RMSE)`, MAE `r fmt(locked_row$Holdout_MAE)`, and
$R^2$ `r fmt(locked_row$Holdout_R2)`. The results support regularization as a
stability device for correlated chemistry measurements, while the limited and
discrete sensory scale constrains attainable accuracy and interpretation.
\end{abstract}

```{r child-sections}
section_files <- file.path(
  REPORT_DIR, "sections",
  c(
    "01_exploratory_data_analysis.Rmd", "02_data_cleaning.Rmd",
    "03_feature_selection.Rmd", "04_regularized_models.Rmd",
    "05_evaluation_conclusion.Rmd", "06_appendix.Rmd"
  )
)
for (section in section_files) {
  cat(knitr::knit_child(section, envir = knitr::knit_global(), quiet = TRUE),
      sep = "\n")
}
```

\newpage

# References {-}
