---
title: "Can Chemistry Predict Red Wine Quality?"
subtitle: "STAT452 Final Project — Part 1"
author: "Group 08"
date: "Academic Year 2025–2026"
output:
  beamer_presentation:
    latex_engine: xelatex
    theme: Madrid
    colortheme: default
    fonttheme: professionalfonts
    slide_level: 2
    keep_tex: false
    includes:
      in_header: header.tex
classoption: aspectratio=169
---

```{r setup, include=FALSE}
input_file <- normalizePath(knitr::current_input(dir = TRUE), winslash = "/",
                            mustWork = TRUE)
source(file.path(dirname(input_file), "..", "setup.R"))
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
knitr::opts_knit$set(root.dir = PART1_ROOT)
options(knitr.graphics.rel_path = FALSE)
if (!file.exists(file.path(OUTPUT_DIR, "analysis_summary.RData"))) {
  source(file.path(PART1_ROOT, "R", "00_run_all.R"),
         local = new.env(parent = globalenv()))
}
load(file.path(OUTPUT_DIR, "analysis_summary.RData"))
performance <- analysis_summary$holdout$performance
locked_model <- analysis_summary$model_lock$Locked_model[[1]]
locked_row <- performance[performance$Model == locked_model, ]
winner <- performance[which.min(performance$Holdout_RMSE), ]
selected <- subset(analysis_summary$lasso_selection, Selected == "Yes")
fmt <- function(x, d = 3) formatC(x, digits = d, format = "f")
pretty_feature <- function(x) gsub("_", " ", x)
```

## Question and design

**Question:** predict expert red-wine quality from 11 physicochemical measurements.

- Regression preserves the ordered 3--8 score without inventing a threshold.
- `r analysis_summary$config$raw_rows` raw rows; `r analysis_summary$config$duplicate_rows`
  exact duplicates removed before splitting.
- Fixed, quality-stratified split:
  `r analysis_summary$config$training_rows` train / `r analysis_summary$config$holdout_rows` holdout.
- Holdout outcomes opened once, after model choice.

**Primary metric:** RMSE; **supporting:** MAE and $R^2$.

## What the training data show

```{r quality-figure, out.width='82%', fig.align='center'}
knitr::include_graphics(file.path(FIGURE_DIR, "fig_p2_quality_distribution.pdf"))
```

Scores 5--6 dominate; rare extremes will be hard to learn precisely.

## Chemistry signals are correlated

```{r correlation-figure, out.width='72%', fig.align='center'}
knitr::include_graphics(file.path(FIGURE_DIR, "fig_p2_correlation_heatmap.pdf"))
```

Alcohol is the strongest positive marginal signal; volatile acidity is the
strongest negative signal. Correlated acidity/density/pH measures motivate
regularization.

## Cleaning and leakage controls

- No missing or non-finite values; median imputation remains a safe no-op.
- IQR rules flag `r analysis_summary$rows_with_any_iqr_flag` training rows, so
  deleting every flag would change the population.
- Winsorize predictors at training 1st/99th percentiles; never clip quality.
- `log1p`: `r paste(pretty_feature(analysis_summary$log_features), collapse = ", ")`.
- Standardize after transformation.
- Every CV fold refits imputation, limits, centering, and scaling.

## Feature relevance

```{r screening-figure, out.width='88%', fig.align='center'}
knitr::include_graphics(file.path(FIGURE_DIR, "fig_p3_feature_screening.pdf"))
```

Lasso one-SE selection retains **`r nrow(selected)` / 11** predictors:
`r paste(pretty_feature(selected$Feature), collapse = ", ")`.

## Models and tuning

We compare mean and OLS baselines with:

\[
\text{loss}+\lambda\{\alpha\|\beta\|_1+(1-\alpha)\|\beta\|_2^2/2\}.
\]

- Ridge: $\alpha=0$
- Lasso: $\alpha=1$
- Elastic Net: CV chose $\alpha=`r fmt(analysis_summary$enet_alpha, 2)`$
- Same five folds and 120 $\lambda$ values for fair comparison

```{r cv-figure, out.width='88%', fig.align='center'}
knitr::include_graphics(file.path(FIGURE_DIR, "fig_p4_cv_curves.pdf"))
```

## Coefficient shrinkage

```{r paths-figure, out.width='90%', fig.align='center'}
knitr::include_graphics(file.path(FIGURE_DIR, "fig_p4_coefficient_paths.pdf"))
```

- Ridge shares weight across correlated variables.
- Lasso yields exact zeros and a sparse interpretation.
- Elastic Net combines grouped shrinkage and selection.

**Locked before holdout:** `r locked_model`, CV RMSE `r fmt(analysis_summary$model_lock$CV_RMSE[[1]])`.

## Untouched holdout results

```{r results-table, results='asis'}
slide_table <- performance[, c("Model", "CV_RMSE", "Holdout_RMSE",
                               "Holdout_MAE", "Holdout_R2", "Locked_by_CV")]
names(slide_table) <- c("Model", "CV RMSE", "Test RMSE", "Test MAE", "$R^2$", "Locked")
cat(knitr::kable(slide_table, format = "latex", booktabs = TRUE, digits = 3,
                 escape = FALSE))
```

Locked `r locked_model`: RMSE **`r fmt(locked_row$Holdout_RMSE)`**, MAE
**`r fmt(locked_row$Holdout_MAE)`**, $R^2$ **`r fmt(locked_row$Holdout_R2)`**.
Observed lowest RMSE: `r winner$Model`; we do not relabel it as the confirmatory
winner after seeing the holdout.

## What the errors mean

```{r predictions-figure, out.width='69%', fig.align='center'}
knitr::include_graphics(file.path(FIGURE_DIR, "fig_p5_actual_vs_predicted.pdf"))
```

All models shrink predictions toward 5--6 and miss rare extremes. Small ranking
differences are less important than the shared stability gain and honest
uncertainty.

## Takeaways and next steps

1. A leakage-controlled regularized workflow improves on the mean benchmark.
2. `r locked_model` is the predeclared recommendation for expected-score prediction.
3. Alcohol and volatile acidity are useful, but no measurement acts alone.
4. Limitations: ordinal target, rare tail scores, unknown duplicate provenance,
   and omitted production/market variables.
5. Next: compare ordinal or robust nonlinear models under the same locked split.

**Reproducibility:** one build command, fixed seeds, modular five-person
workstreams, PDF figures, CSV/LaTeX tables, model lock, artifact checksums, and
session information.
