% =============================================================================
% 03_math_mechanisms.tex — Problem 3: Mathematical Mechanisms (P4 — Triết)
% =============================================================================
\section{Mathematical Mechanisms}
\label{sec:math}

\subsection{Ridge Regression and Conditioning}
\label{sec:ridge-math}

Let $\bX\in\R^{n\times p}$ be the centred and scaled training design, let
$\by\in\R^n$ be the centred training response, and define
$G=\bX^\top\bX/n$ and $\bm{z}=\bX^\top\by/n$.  The intercept is handled
separately and is not penalised.  Under the same scaling used by
\texttt{glmnet}, the Ridge objective is
\begin{equation}
\label{eq:ridge-obj}
 Q_\lambda(\bbeta)=
 \frac{1}{2n}\norm{\by-\bX\bbeta}_2^2
 +\frac{\lambda}{2}\norm{\bbeta}_2^2,
 \qquad \lambda>0.
\end{equation}
Its gradient is
\begin{align}
 \nabla Q_\lambda(\bbeta)
 &= -\frac{1}{n}\bX^\top(\by-\bX\bbeta)+\lambda\bbeta \\
 &= G\bbeta-\bm{z}+\lambda\bbeta.
\end{align}
Setting this equal to zero gives the normal equations and solution
\begin{equation}
 (G+\lambda\bm{I}_p)\hat\bbeta=\bm{z},
 \qquad
 \boxed{\hat\bbeta^{\mathrm{Ridge}}_\lambda
 =(G+\lambda\bm{I}_p)^{-1}\bm{z}}.
\label{eq:ridge-solution}
\end{equation}
Because $G$ is positive semidefinite, $G+\lambda I_p$ is positive definite
for every $\lambda>0$, so this minimiser is unique.

If $\mu_{\max}\geq\cdots\geq\mu_{\min}\geq0$ are the eigenvalues of $G$,
adding Ridge translates every eigenvalue by $\lambda$.  Hence
\begin{equation}
 \kappa_2(G+\lambda\bm I_p)
 =\frac{\mu_{\max}+\lambda}{\mu_{\min}+\lambda}
 \leq \kappa_2(G).
\label{eq:ridge-condition}
\end{equation}
Geometrically, the objective gains curvature in every eigendirection;
numerically, its flattest direction is moved away from zero.

For the training design, $\mu_{\max}=8.942970$ and
$\mu_{\min}=0.002824$, so $\kappa_2(G)=3166.75$.  At the fitted
$\lambda_{\min}=0.083342$,
\[
 \frac{8.942970+0.083342}{0.002824+0.083342}=104.755,
\]
matching Table~\ref{tab:condition_numbers}; the condition number falls by a
factor of 30.2.

\subsection{Lasso Optimality and Soft Thresholding}
\label{sec:lasso-math}

The Lasso objective on the same scale is
\begin{equation}
 R_\lambda(\bbeta)=
 \frac{1}{2n}\norm{\by-\bX\bbeta}_2^2+\lambda\norm{\bbeta}_1.
\label{eq:lasso-obj}
\end{equation}
Because $|\beta_j|$ is not differentiable at zero, coordinate $j$ satisfies
\begin{equation}
 -\frac{1}{n}\bx_j^\top(\by-\bX\hat\bbeta)
 +\lambda\,\partial|\hat\beta_j|\ni0,
\label{eq:lasso-subgradient}
\end{equation}
where $\partial|b|=\{\mathrm{sign}(b)\}$ for $b\ne0$ and
$\partial|0|=[-1,1]$.  Writing $r=\by-\bX\hat\bbeta$, this is equivalent to
\[
\begin{cases}
 n^{-1}\bx_j^\top r=\lambda\,\mathrm{sign}(\hat\beta_j),
     & \hat\beta_j\ne0,\\
 \left|n^{-1}\bx_j^\top r\right|\leq\lambda,
     & \hat\beta_j=0.
\end{cases}
\]

When $G=I_p$, the coordinates decouple.  Completing the square for coordinate
$j$ gives the soft-thresholding rule
\begin{equation}
 \boxed{\hat\beta_j
 =\mathrm{sign}(z_j)(|z_j|-\lambda)_+}.
\label{eq:soft-threshold}
\end{equation}
Thus $|z_j|\leq\lambda$ produces an exact zero, while larger values are shrunk
toward zero by $\lambda$.  The $\ell_1$ constraint has corners on coordinate
axes, so a smooth loss contour can first touch a corner.  The $\ell_2$
constraint is smooth and generally shrinks coefficients without landing
exactly on an axis.

\subsection{Connecting Algebra to Evidence}
\label{sec:algebra-evidence}

The fitted \texttt{chest} slope illustrates the different mechanisms.  OLS
estimates $-0.08$; Ridge at $\lambda_{\min}$ redistributes signal among the
correlated measurements and returns 0.82 but does not select the variable
away; Lasso returns exactly zero.  The Lasso residual correlation for this
coordinate can satisfy the interval condition in
Equation~\ref{eq:lasso-subgradient}, whereas Ridge must satisfy the smooth
system in Equation~\ref{eq:ridge-solution}.  Together with the observed
condition-number reduction, this connects the algebra to the numerical
evidence in Section~\ref{sec:ols-ridge-lasso}.
