---
title: "ASESII 24A02 Project-Based Quiz 1"
subtitle: "Ridge, Lasso, and Regularization for Neural Features"
author:
  - "Group 01"
  - "Nguyễn Đình Thiên Lộc (24125093), Trần Lê Anh Tuấn (24125107)"
  - "Lê Minh Thuận (24125105), Nguyễn Bảo Minh Triết (24125047)"
  - "Nguyễn Hồng Tấn Tài (24125078)"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
    fig_crop: false
  html_document:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
header-includes:
  - '\renewcommand{\and}{\\}'
fontsize: 11pt
geometry: margin=1in
---

# Authorship and contribution statement {-}

We confirm that every group member can explain the submitted analysis. AI use is reported in Section 5C, and the group checked every AI-assisted item used in this report.

| Student | ID | Main contributions | Verification performed |
|---|---|---|---|
| Nguyễn Đình Thiên Lộc | 24125093 | P1 — project configuration, data preprocessing, exploratory data analysis, data leakage prevention, and drafting the Prediction Design section. | Verified the exclusion of linearly dependent predictors and confirmed no test-set information leaked into the standardisation pipeline. |
| Trần Lê Anh Tuấn | 24125107 | P2 — OLS baseline, Ridge estimation, cross-validation curves, coefficient paths, and conditioning diagnostics. | Checked the shared-fold OLS and Ridge results, both tuning rules, generated tables, and diagnostic figures. |
| Lê Minh Thuận | 24125105 | P2 — Lasso estimation, controlled model comparison, core-model declaration, and bootstrap stability analysis. | Confirmed the Lasso active sets, shared-fold errors, pre-holdout selection rule, and bootstrap summaries. |
| Nguyễn Bảo Minh Triết | 24125047 | P3 — Ridge normal-equation and conditioning derivations, Lasso subgradient conditions, and evidence connection. | Re-derived the stated results under the same `glmnet` scaling and checked them against the numerical eigenvalues. |
| Nguyễn Hồng Tấn Tài | 24125078 | P4--P5 — Elastic Net, fixed ReLU features, locked holdout evaluation, integration, abstract, and reproducibility record. | Verified the frozen feature map, training-only hidden-feature handling, locked declaration, final metrics, and report build. |

# Abstract {-}

We predict Brozek body-fat percentage from 14 anthropometric measurements for
252 adult men.  Outcome-derived variables were excluded, preprocessing used
training-only statistics, and OLS, Ridge, Lasso, Elastic Net, and penalised
models on 200 fixed random ReLU features were compared with one shared
five-fold split.  Lasso at $\lambda_{\min}=0.0936$ had the lowest training
cross-validated RMSE among the original-predictor models (4.242), retained nine
slopes, and was declared for deployment before the holdout responses were
examined.  On the 51-subject holdout set it remained best, with RMSE 4.247,
MAE 3.524, and $R^2=0.684$.  Elastic Net was close (RMSE 4.267), whereas the
fixed ReLU expansions did not improve prediction.  We therefore recommend the
predeclared Lasso for this population, while treating its selected variables as
predictive rather than causal evidence.

# Reproducible setup {-}

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c("glmnet", "knitr")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  stop("Install the required packages before rendering: ",
       paste(missing_packages, collapse = ", "))
}

library(glmnet)
library(knitr)

seeds <- list(
  split = 240201L,
  cv = 240301L,
  features = 240401L
)
```

## Shared helper functions {-}

```{r helpers}
find_exam_data <- function(filename) {
  input <- knitr::current_input(dir = TRUE)
  input_dir <- if (length(input) && nzchar(input)) dirname(input) else getwd()
  candidates <- c(
    file.path(input_dir, "data", filename),
    file.path(input_dir, "..", "data", filename),
    file.path(getwd(), "data", filename)
  )
  existing <- candidates[file.exists(candidates)]
  if (!length(existing)) stop("Cannot locate ", filename, " by a relative path.")
  hits <- unique(normalizePath(existing, winslash = "/", mustWork = TRUE))
  if (length(hits) > 1L && length(unique(unname(tools::md5sum(hits)))) > 1L) {
    stop("Multiple nonidentical copies of ", filename, " were found.")
  }
  hits[[1L]]
}

split_rows <- function(n, train_prop = 0.80, seed) {
  stopifnot(n >= 10L, train_prop > 0, train_prop < 1)
  set.seed(seed)
  train <- sort(sample.int(n, floor(train_prop * n), replace = FALSE))
  list(train = train, test = setdiff(seq_len(n), train))
}

fit_scaler <- function(x, tol = 1e-12) {
  x <- as.matrix(x)
  center <- colMeans(x)
  scale <- apply(x, 2L, sd)
  keep <- is.finite(scale) & scale > tol
  if (!any(keep)) stop("No nonconstant training predictor remains.")
  list(
    center = center[keep],
    scale = scale[keep],
    keep = keep,
    dropped = colnames(x)[!keep]
  )
}

apply_scaler <- function(x, prep) {
  x <- as.matrix(x)[, prep$keep, drop = FALSE]
  scaled <- scale(x, center = prep$center, scale = prep$scale)
  attr(scaled, "scaled:center") <- NULL
  attr(scaled, "scaled:scale") <- NULL
  scaled
}

make_foldid <- function(n, k = 5L, seed) {
  if (k < 3L || k > n) stop("Require 3 <= k <= number of training rows.")
  set.seed(seed)
  sample(rep(seq_len(k), length.out = n))
}

default_lambda_grid <- function() {
  exp(seq(log(1e-4), log(1e3), length.out = 140L))
}

foldclean_select_lambda <- function(fold_mse, lambda) {
  cvm <- colMeans(fold_mse)
  cvsd <- apply(fold_mse, 2, sd) / sqrt(nrow(fold_mse))
  min_idx <- which.min(cvm)
  threshold <- cvm[min_idx] + cvsd[min_idx]
  eligible <- which(cvm <= threshold)
  one_se_idx <- eligible[which.max(lambda[eligible])]
  list(
    cvm = cvm,
    cvsd = cvsd,
    lambda.min = lambda[min_idx],
    lambda.1se = lambda[one_se_idx]
  )
}

fit_foldclean_glmnet <- function(x_raw, y, alpha, foldid,
                                 lambda = default_lambda_grid()) {
  x_raw <- as.matrix(x_raw)
  y <- as.numeric(y)
  lambda <- sort(lambda, decreasing = TRUE)
  folds <- sort(unique(foldid))
  fold_mse <- matrix(NA_real_, nrow = length(folds), ncol = length(lambda))

  for (i in seq_along(folds)) {
    val_idx <- which(foldid == folds[i])
    train_idx <- which(foldid != folds[i])
    fold_scaler <- fit_scaler(x_raw[train_idx, , drop = FALSE])
    x_fold_train <- apply_scaler(x_raw[train_idx, , drop = FALSE], fold_scaler)
    x_fold_val <- apply_scaler(x_raw[val_idx, , drop = FALSE], fold_scaler)
    fold_fit <- glmnet::glmnet(
      x = x_fold_train,
      y = y[train_idx],
      family = "gaussian",
      alpha = alpha,
      lambda = lambda,
      standardize = FALSE,
      intercept = TRUE
    )
    pred <- predict(fold_fit, newx = x_fold_val, s = lambda)
    fold_mse[i, ] <- colMeans((y[val_idx] - pred)^2)
  }

  selected <- foldclean_select_lambda(fold_mse, lambda)
  final_scaler <- fit_scaler(x_raw)
  x_final <- apply_scaler(x_raw, final_scaler)
  final_fit <- glmnet::glmnet(
    x = x_final,
    y = y,
    family = "gaussian",
    alpha = alpha,
    lambda = lambda,
    standardize = FALSE,
    intercept = TRUE
  )

  fit <- list(
    alpha = alpha,
    lambda = lambda,
    cvm = selected$cvm,
    cvsd = selected$cvsd,
    lambda.min = selected$lambda.min,
    lambda.1se = selected$lambda.1se,
    fold_mse = fold_mse,
    scaler = final_scaler,
    glmnet.fit = final_fit,
    transform = "scaled"
  )
  class(fit) <- "foldclean_glmnet"
  fit
}

fit_foldclean_neural_glmnet <- function(x_raw, y, foldid, A, bias, alpha,
                                        lambda = default_lambda_grid()) {
  x_raw <- as.matrix(x_raw)
  y <- as.numeric(y)
  lambda <- sort(lambda, decreasing = TRUE)
  folds <- sort(unique(foldid))
  fold_mse <- matrix(NA_real_, nrow = length(folds), ncol = length(lambda))
  relu <- function(z) pmax(z, 0)

  for (i in seq_along(folds)) {
    val_idx <- which(foldid == folds[i])
    train_idx <- which(foldid != folds[i])
    x_fold_scaler <- fit_scaler(x_raw[train_idx, , drop = FALSE])
    x_fold_train <- apply_scaler(x_raw[train_idx, , drop = FALSE], x_fold_scaler)
    x_fold_val <- apply_scaler(x_raw[val_idx, , drop = FALSE], x_fold_scaler)
    h_train_raw <- relu(x_fold_train %*% A +
      matrix(bias, nrow = nrow(x_fold_train), ncol = ncol(A), byrow = TRUE))
    h_val_raw <- relu(x_fold_val %*% A +
      matrix(bias, nrow = nrow(x_fold_val), ncol = ncol(A), byrow = TRUE))
    h_fold_scaler <- fit_scaler(h_train_raw)
    h_fold_train <- apply_scaler(h_train_raw, h_fold_scaler)
    h_fold_val <- apply_scaler(h_val_raw, h_fold_scaler)
    fold_fit <- glmnet::glmnet(
      x = h_fold_train,
      y = y[train_idx],
      family = "gaussian",
      alpha = alpha,
      lambda = lambda,
      standardize = FALSE,
      intercept = TRUE
    )
    pred <- predict(fold_fit, newx = h_fold_val, s = lambda)
    fold_mse[i, ] <- colMeans((y[val_idx] - pred)^2)
  }

  selected <- foldclean_select_lambda(fold_mse, lambda)
  x_final_scaler <- fit_scaler(x_raw)
  x_final <- apply_scaler(x_raw, x_final_scaler)
  h_final_raw <- relu(x_final %*% A +
    matrix(bias, nrow = nrow(x_final), ncol = ncol(A), byrow = TRUE))
  h_final_scaler <- fit_scaler(h_final_raw)
  h_final <- apply_scaler(h_final_raw, h_final_scaler)
  final_fit <- glmnet::glmnet(
    x = h_final,
    y = y,
    family = "gaussian",
    alpha = alpha,
    lambda = lambda,
    standardize = FALSE,
    intercept = TRUE
  )

  fit <- list(
    alpha = alpha,
    lambda = lambda,
    cvm = selected$cvm,
    cvsd = selected$cvsd,
    lambda.min = selected$lambda.min,
    lambda.1se = selected$lambda.1se,
    fold_mse = fold_mse,
    x_scaler = x_final_scaler,
    h_scaler = h_final_scaler,
    A = A,
    bias = bias,
    glmnet.fit = final_fit,
    transform = "neural"
  )
  class(fit) <- "foldclean_glmnet"
  fit
}

coef.foldclean_glmnet <- function(object, s = "lambda.min", ...) {
  s_value <- if (is.character(s)) object[[s]] else s
  coef(object$glmnet.fit, s = s_value, ...)
}

predict.foldclean_glmnet <- function(object, newx, s = "lambda.min", ...) {
  s_value <- if (is.character(s)) object[[s]] else s
  newx <- as.matrix(newx)
  if (identical(object$transform, "scaled")) {
    transformed <- apply_scaler(newx, object$scaler)
  } else {
    relu <- function(z) pmax(z, 0)
    x_scaled <- apply_scaler(newx, object$x_scaler)
    h_raw <- relu(x_scaled %*% object$A +
      matrix(object$bias, nrow = nrow(x_scaled),
             ncol = ncol(object$A), byrow = TRUE))
    transformed <- apply_scaler(h_raw, object$h_scaler)
  }
  predict(object$glmnet.fit, newx = transformed, s = s_value, ...)
}

plot.foldclean_glmnet <- function(x, main = "Fold-clean CV Error", ...) {
  upper <- x$cvm + x$cvsd
  lower <- x$cvm - x$cvsd
  plot(log(x$lambda), x$cvm, type = "b", pch = 16,
       ylim = range(c(lower, upper), finite = TRUE),
       xlab = expression(log(lambda)), ylab = "Cross-Validation MSE",
       main = main, ...)
  arrows(log(x$lambda), lower, log(x$lambda), upper,
         angle = 90, code = 3, length = 0.03, col = "grey60")
  abline(v = log(x$lambda.min), lty = 2, col = "red")
  abline(v = log(x$lambda.1se), lty = 2, col = "blue")
}

score_regression <- function(y, prediction) {
  y <- as.numeric(y)
  prediction <- as.numeric(prediction)
  if (length(y) != length(prediction) || any(!is.finite(c(y, prediction)))) {
    stop("Metrics require equal-length finite vectors.")
  }
  residuals <- y - prediction
  rss <- sum(residuals^2)
  tss <- sum((y - mean(y))^2)
  r2 <- 1 - rss/tss
  c(RMSE = sqrt(mean(residuals^2)),
    MAE = mean(abs(residuals)),
    R2 = r2)
}

count_nonzero <- function(fit, s = "lambda.min", tol = 1e-8) {
  b <- as.matrix(stats::coef(fit, s = s))
  slopes <- b[rownames(b) != "(Intercept)", 1L]
  sum(abs(slopes) > tol)
}

safe_condition_numbers <- function(x, lambda, tol = 1e-10) {
  eigenvalues <- eigen(
    crossprod(x) / nrow(x),
    symmetric = TRUE,
    only.values = TRUE
  )$values
  largest <- max(eigenvalues)
  smallest <- max(min(eigenvalues), 0)
  cutoff <- tol * max(1, largest)
  c(
    gram = if (smallest <= cutoff) Inf else largest / smallest,
    ridge = (largest + lambda) / (smallest + lambda)
  )
}
```

# Problem 1. Prediction design and leakage control

## 1A. Target and predictors

```{r data-configuration}
config <- list(
  file = "fat.csv",
  response = "brozek",
  excluded = c("brozek", "siri", "density", "free")
)

data_path <- find_exam_data(config$file)
data_raw <- read.csv(data_path, check.names = FALSE)
predictors <- setdiff(names(data_raw), config$excluded)
analysis_data <- data_raw[, c(config$response, predictors), drop = FALSE]

if (anyNA(analysis_data)) stop("Declare a training-only missing-value plan.")
if (!all(vapply(analysis_data, is.numeric, logical(1)))) {
  stop("The assigned response and predictors must be numeric.")
}
```

The units of observation are adult male subjects. The response variable is `brozek` (body fat percentage derived from density). Predictors `siri`, `density`, and `free` are excluded to prevent data leakage because they are algebraic transformations or near-perfect proxies of the body fat calculation, which would make prediction trivially deterministic or circular. This leaves 14 anthropometric measurements as candidate predictors.

## 1B. Split and preprocessing plan

```{r split-and-scale}
split <- split_rows(nrow(analysis_data), train_prop = 0.80, seed = seeds$split)
train_id <- split$train
test_id <- split$test

x_raw <- as.matrix(analysis_data[, predictors, drop = FALSE])
y_raw <- analysis_data[[config$response]]

x_raw_train <- x_raw[train_id, , drop = FALSE]
x_raw_test <- x_raw[test_id, , drop = FALSE]
x_prep <- fit_scaler(x_raw_train)
x_train <- apply_scaler(x_raw_train, x_prep)
x_test <- apply_scaler(x_raw_test, x_prep)
y_train <- y_raw[train_id]
y_test <- y_raw[test_id]

foldid <- make_foldid(nrow(x_train), k = 5L, seed = seeds$cv)
```

The data split uses seed `240201`, yielding $n_{\mathrm{train}} = 201$ and $n_{\mathrm{test}} = 51$. Center/scale parameters are computed only from the training set and applied to the test set, preventing leakage. K=5 cross-validation folds are assigned using seed `240301` and reused across all models for consistency.

## 1C. Training-data exploration

```{r training-eda, fig.width=8, fig.height=6}
# Response distribution
hist(y_train, breaks = 15, col = "gray80", border = "white",
     main = "Distribution of brozek (Training Set)", xlab = "Body Fat % (brozek)")

# Heatmap
cor_mat <- cor(analysis_data[train_id, ])
image(1:ncol(cor_mat), 1:ncol(cor_mat), cor_mat[, ncol(cor_mat):1],
      col = colorRampPalette(c("#2166ac", "#f7f7f7", "#b2182b"))(100),
      axes = FALSE, main = "Correlation Heatmap (Training Set)", xlab = "", ylab = "")
axis(1, at = 1:ncol(cor_mat), labels = colnames(cor_mat), las = 2, cex.axis = 0.7)
axis(2, at = 1:ncol(cor_mat), labels = rev(colnames(cor_mat)), las = 2, cex.axis = 0.7)

# Anomaly check
boxplot(x_train, las = 2, main = "Standardised Predictors (Training Set)")
```

The training response `brozek` is roughly symmetric and bell-shaped. Multicollinearity is severe among size variables, such as `hip` and `weight` ($r = 0.945$) and `chest` and `weight` ($r = 0.894$). One notable anomaly is original row 39, which exhibits `hip` = 147.7, over 6.7 training standard deviations above the mean. We ask: can regularised regression improve prediction stability and identify which measurements are most useful for predicting body fat?

# Problem 2. OLS, ridge, and lasso

## 2A. OLS baseline

```{r ols, fig.width=8, fig.height=4}
ols_fit <- lm(y ~ ., data = data.frame(y = y_train, x_train))
y_pred_ols_train <- predict(ols_fit, newdata = data.frame(x_train))
ols_train_scores <- score_regression(y_train, y_pred_ols_train)

# K-fold CV for OLS
ols_cv_pred <- rep(NA_real_, length(y_train))
for (i in 1:5) {
  val_idx <- which(foldid == i)
  trn_idx <- which(foldid != i)
  fold_scaler <- fit_scaler(x_raw_train[trn_idx, , drop = FALSE])
  x_fold_train <- apply_scaler(x_raw_train[trn_idx, , drop = FALSE], fold_scaler)
  x_fold_val <- apply_scaler(x_raw_train[val_idx, , drop = FALSE], fold_scaler)
  fold_fit <- lm(y ~ ., data = data.frame(y = y_train[trn_idx], x_fold_train))
  ols_cv_pred[val_idx] <- predict(fold_fit, newdata = data.frame(x_fold_val))
}
ols_cv_rmse <- sqrt(mean((y_train - ols_cv_pred)^2))

# Diagnostic Plot
plot(fitted(ols_fit), residuals(ols_fit), pch = 16, col = "gray40",
     main = "OLS: Residuals vs Fitted", xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, lty = 2, col = "red")
lines(lowess(fitted(ols_fit), residuals(ols_fit)), col = "blue", lwd = 2)

cat("OLS Train RMSE:", ols_train_scores[["RMSE"]], "\n")
cat("OLS CV RMSE:", ols_cv_rmse, "\n")
```

Under OLS, the partial slopes for highly collinear predictors are large and counterintuitive (e.g. `abdom` is 9.47, whereas `weight`, `adipos`, and `hip` are negative). OLS yields a training RMSE of 3.884 and a CV RMSE of 4.406. Residual and Q-Q plots suggest tail departures from Gaussian assumptions but show no severe non-linearity.

## 2B. Ridge

```{r ridge, fig.width=8, fig.height=4}
ridge_fit <- fit_foldclean_glmnet(x_raw_train, y_train, alpha = 0, foldid = foldid)

# Plot CV curve
plot(ridge_fit, main = "Ridge Cross-Validation Curve")

# Plot coefficient path
ridge_path <- glmnet(x_train, y_train, alpha = 0, standardize = FALSE)
plot(ridge_path, xvar = "lambda", main = "Ridge Coefficient Path")

cat("Ridge lambda.min:", ridge_fit$lambda.min, "\n")
cat("Ridge lambda.1se:", ridge_fit$lambda.1se, "\n")
```

Ridge selects $\lambda_{\min} = `r sprintf("%.4f", ridge_fit$lambda.min)`$ and $\lambda_{1\mathrm{se}} = `r sprintf("%.4f", ridge_fit$lambda.1se)`$. It shrinks all slopes toward zero but retains all `r count_nonzero(ridge_fit)` variables. The `abdom` slope changes from `r sprintf("%.2f", coef(ols_fit)["abdom"])` under OLS to `r sprintf("%.2f", as.matrix(coef(ridge_fit, s = "lambda.min"))["abdom", 1])` at $\lambda_{\min}$ and `r sprintf("%.2f", as.matrix(coef(ridge_fit, s = "lambda.1se"))["abdom", 1])` at $\lambda_{1\mathrm{se}}$. Adding the Ridge penalty reduces the condition number from `r sprintf("%.2f", safe_condition_numbers(x_train, ridge_fit$lambda.min)["gram"])` to `r sprintf("%.2f", safe_condition_numbers(x_train, ridge_fit$lambda.min)["ridge"])`.

## 2C. Lasso

```{r lasso, fig.width=8, fig.height=4}
lasso_fit <- fit_foldclean_glmnet(x_raw_train, y_train, alpha = 1, foldid = foldid)

# Plot CV curve
plot(lasso_fit, main = "Lasso Cross-Validation Curve")

# Plot coefficient path
lasso_path <- glmnet(x_train, y_train, alpha = 1, standardize = FALSE)
plot(lasso_path, xvar = "lambda", main = "Lasso Coefficient Path")

cat("Lasso lambda.min:", lasso_fit$lambda.min, "\n")
cat("Lasso lambda.1se:", lasso_fit$lambda.1se, "\n")
```

At $\lambda_{\min} = `r sprintf("%.4f", lasso_fit$lambda.min)`$, five-fold cross-validation selects `r count_nonzero(lasso_fit)` active predictors (`r paste(rownames(as.matrix(coef(lasso_fit, s = "lambda.min")))[-1][abs(as.matrix(coef(lasso_fit, s = "lambda.min"))[-1, 1]) > 1e-8], collapse = ", ")`) with CV RMSE `r sprintf("%.3f", sqrt(min(lasso_fit$cvm)))`. At $\lambda_{1\mathrm{se}} = `r sprintf("%.4f", lasso_fit$lambda.1se)`$, `r count_nonzero(lasso_fit, s = "lambda.1se")` predictors remain active.

## 2D. Training-based comparison and locked core model

```{r core-comparison}
# Table construction
ols_coef <- coef(ols_fit)[-1]
ols_norm_l1 <- sum(abs(ols_coef))
ols_norm_l2 <- sqrt(sum(ols_coef^2))

ridge_coef <- as.matrix(coef(ridge_fit, s = "lambda.min"))[-1, 1]
ridge_norm_l1 <- sum(abs(ridge_coef))
ridge_norm_l2 <- sqrt(sum(ridge_coef^2))

lasso_coef <- as.matrix(coef(lasso_fit, s = "lambda.min"))[-1, 1]
lasso_norm_l1 <- sum(abs(lasso_coef))
lasso_norm_l2 <- sqrt(sum(lasso_coef^2))

comparison_df <- data.frame(
  Model = c("OLS", "Ridge (lambda.min)", "Lasso (lambda.min)"),
  Tuning = c("None", paste0("lambda=", round(ridge_fit$lambda.min, 4)), paste0("lambda=", round(lasso_fit$lambda.min, 4))),
  CV_RMSE = c(ols_cv_rmse, sqrt(ridge_fit$cvm[ridge_fit$lambda == ridge_fit$lambda.min]), sqrt(lasso_fit$cvm[lasso_fit$lambda == lasso_fit$lambda.min])),
  L1_Norm = c(ols_norm_l1, ridge_norm_l1, lasso_norm_l1),
  L2_Norm = c(ols_norm_l2, ridge_norm_l2, lasso_norm_l2),
  Nonzero = c(14, 14, sum(abs(lasso_coef) > 1e-8))
)
kable(comparison_df, digits = 4, caption = "Training-based Model Comparison")
```

**Core model nominated before holdout evaluation:** Lasso at $\lambda_{\min} = `r sprintf("%.4f", lasso_fit$lambda.min)`$. It achieves the lowest cross-validated RMSE (`r sprintf("%.3f", min(comparison_df$CV_RMSE))`) while selecting `r count_nonzero(lasso_fit)` predictors.

## 2E. Perturbation or stability check

**Prediction before running the check:** We predict that Ridge coefficients will remain stable under perturbation due to the $L_2$ penalty, while Lasso will select different variables among highly correlated sets across resamples.

```{r perturbation, fig.width=8, fig.height=4}
set.seed(240302L)
B <- 50
lasso_sel <- matrix(0, nrow = B, ncol = ncol(x_train))
ridge_coefs_boot <- matrix(0, nrow = B, ncol = ncol(x_train))
ols_coefs_boot <- matrix(0, nrow = B, ncol = ncol(x_train))
colnames(lasso_sel) <- colnames(x_train)
colnames(ridge_coefs_boot) <- colnames(x_train)
colnames(ols_coefs_boot) <- colnames(x_train)

for (b in 1:B) {
  boot_idx <- sample(nrow(x_train), replace = TRUE)
  x_b <- x_train[boot_idx, ]
  y_b <- y_train[boot_idx]
  
  # Fits
  fit_ols_b <- lm(y ~ ., data = data.frame(y = y_b, x_b))
  ols_coefs_boot[b, ] <- coef(fit_ols_b)[-1]
  
  fit_ridge_b <- glmnet(x_b, y_b, alpha = 0, lambda = ridge_fit$lambda.min, standardize = FALSE)
  ridge_coefs_boot[b, ] <- as.matrix(coef(fit_ridge_b))[-1, 1]
  
  fit_lasso_b <- glmnet(x_b, y_b, alpha = 1, lambda = lasso_fit$lambda.min, standardize = FALSE)
  lasso_coef_b <- as.matrix(coef(fit_lasso_b))[-1, 1]
  lasso_sel[b, ] <- abs(lasso_coef_b) > 1e-8
}

# Selection frequencies
freqs <- colMeans(lasso_sel)
barplot(sort(freqs, decreasing = TRUE), las = 2, main = "Lasso Variable Selection Frequency (B=50)")
```

**Observed result and explanation:** Lasso selects `abdom` in `r sprintf("%.0f%%", 100 * freqs["abdom"])` and `wrist` in `r sprintf("%.0f%%", 100 * freqs["wrist"])` of bootstrap resamples, while correlated size predictors vary more. For `weight`, the bootstrap coefficient SD is `r sprintf("%.2f", sd(ridge_coefs_boot[, "weight"]))` under Ridge versus `r sprintf("%.2f", sd(ols_coefs_boot[, "weight"]))` under OLS, illustrating Ridge's stabilising effect.

# Problem 3. Mathematical mechanisms

## 3A. Ridge and conditioning

Starting from:
$$Q_\lambda(\mathbf{b}) = \frac{1}{2n}\left\lVert \mathbf{y} - \mathbf{Xb} \right\rVert_2^2 + \frac{\lambda}{2}\left\lVert \mathbf{b} \right\rVert_2^2$$

The gradient is:
$$\nabla Q_\lambda(\mathbf{b}) = -\frac{1}{n}\mathbf{X}^\top(\mathbf{y} - \mathbf{Xb}) + \lambda \mathbf{b} = \mathbf{Gb} - \mathbf{z} + \lambda \mathbf{b}$$
where $\mathbf{G} = \frac{1}{n}\mathbf{X}^\top\mathbf{X}$ and $\mathbf{z} = \frac{1}{n}\mathbf{X}^\top\mathbf{y}$. Setting the gradient to zero yields the normal equations:
$$(\mathbf{G} + \lambda \mathbf{I}_p)\mathbf{b} = \mathbf{z}$$
Since $\mathbf{G}$ is positive semidefinite, for any $\lambda > 0$, $(\mathbf{G} + \lambda \mathbf{I}_p)$ is positive definite and invertible, giving the unique minimizer:
$$\hat{\mathbf{b}}_{\mathrm{ridge}} = (\mathbf{G} + \lambda \mathbf{I}_p)^{-1}\mathbf{z}$$

```{r ridge-conditioning}
cond_analysis <- safe_condition_numbers(x_train, lambda = ridge_fit$lambda.min)
cat("Condition number of Gram matrix G:", cond_analysis["gram"], "\n")
cat("Condition number of G + lambda I:", cond_analysis["ridge"], "\n")
```

The condition number drops from `r sprintf("%.2f", cond_analysis["gram"])` to `r sprintf("%.2f", cond_analysis["ridge"])` at $\lambda_{\min} = `r sprintf("%.4f", ridge_fit$lambda.min)`$, a `r sprintf("%.1f", cond_analysis["gram"] / cond_analysis["ridge"])`-fold reduction. Geometrically, adding $\lambda \mathbf{I}$ raises the smallest eigenvalues and prevents division by near-zero values.

## 3B. Lasso optimality and soft thresholding

The Lasso objective is:
$$R_\lambda(\mathbf{b}) = \frac{1}{2n}\left\lVert \mathbf{y} - \mathbf{Xb} \right\rVert_2^2 + \lambda \left\lVert \mathbf{b} \right\rVert_1$$

Since the L1 norm is non-differentiable at $b_j = 0$, we use subgradients. The subgradient condition for coordinate $j$ is:
$$-\frac{1}{n}\mathbf{x}_j^\top(\mathbf{y} - \mathbf{Xb}) + \lambda s_j = 0$$
where $s_j = \mathrm{sign}(b_j)$ if $b_j \ne 0$ and $s_j \in [-1, 1]$ if $b_j = 0$.

When the columns of $\mathbf{X}$ are orthonormal ($\mathbf{G} = \mathbf{I}_p$), the coordinates decouple. For each $j$:
$$-z_j + b_j + \lambda s_j = 0 \implies b_j = z_j - \lambda s_j$$
- If $|z_j| \le \lambda$, then $b_j = 0$ is optimal since $s_j = z_j / \lambda \in [-1, 1]$.
- If $z_j > \lambda$, then $b_j > 0$ and $s_j = 1$, giving $b_j = z_j - \lambda$.
- If $z_j < -\lambda$, then $b_j < 0$ and $s_j = -1$, giving $b_j = z_j + \lambda$.

Combining these yields the soft-thresholding rule:
$$\hat{b}_j = \mathrm{sign}(z_j)(|z_j| - \lambda)_+$$

Because the subgradient condition allows $b_j = 0$ over a range of values (whenever $|z_j| \le \lambda$), Lasso produces exact zeros. In contrast, the Ridge gradient condition is smooth and only yields $b_j = 0$ if the correlation is exactly zero.

## 3C. Connect algebra to evidence

The coefficient for `chest` illustrates this mechanism. OLS estimates a partial slope of `r sprintf("%.2f", coef(ols_fit)["chest"])`. Ridge at $\lambda_{\min}$ estimates `r sprintf("%.2f", as.matrix(coef(ridge_fit, s = "lambda.min"))["chest", 1])`, keeping it active. Lasso sets the coefficient exactly to zero because the conditional correlation falls within the subgradient interval $[-\lambda, \lambda]$.

# Problem 4. Elastic net and a neural-feature bridge

## 4A. Research question and source map

**Research direction:** weight decay, sparse weights, and dropout.

| Claim | Exact primary source and locator | What it establishes | What it does not establish |
|---|---|---|---|
| A quadratic weight penalty yields a decay term in gradient learning. | Krogh and Hertz (1991), Section 1, Eq. (1)-(2). | Shows that L2 penalty adds a $-\lambda w_i$ term directly to standard SGD updates. | Does not analyze adaptive step optimizers (Adam) or demonstrate equivalence there. |
| L2 regularization and weight decay decouple under adaptive updates. | Loshchilov and Hutter (2019), abstract and Section 2. | Shows L2 regularization is not equivalent to weight decay for Adam; motivates AdamW. | Does not make our frozen-feature model equivalent to an end-to-end trained network. |
| Dropout randomly masks units during training and uses the unthinned network at test time. | Srivastava et al. (2014), Section 2 and Fig. 2. | Defines stochastic activation masking during training and scaling of weights at test time. | Does not establish permanent sparsity or pruning of learned weights. |

## 4B. Elastic net on original predictors

```{r original-elastic-net, fig.width=8, fig.height=4}
alpha_grid <- c(0.10, 0.25, 0.50, 0.75, 0.90)
enet_fits <- list()
cv_mses <- numeric(length(alpha_grid))

for (i in seq_along(alpha_grid)) {
  enet_fits[[i]] <- fit_foldclean_glmnet(
    x_raw_train, y_train, alpha = alpha_grid[i], foldid = foldid
  )
  cv_mses[i] <- min(enet_fits[[i]]$cvm)
}

best_idx <- which.min(cv_mses)
best_alpha <- alpha_grid[best_idx]
best_enet <- enet_fits[[best_idx]]

# Plot CV curve
plot(best_enet, main = paste("Elastic Net CV Curve (alpha =", best_alpha, ")"))

# Plot coefficient path
enet_path <- glmnet(x_train, y_train, alpha = best_alpha, standardize = FALSE)
plot(enet_path, xvar = "lambda", main = paste("Elastic Net Coeff Path (alpha =", best_alpha, ")"))

cat("Selected alpha:", best_alpha, "\n")
cat("Selected lambda.min:", best_enet$lambda.min, "\n")
```

The selected Elastic Net model has $\alpha = `r sprintf("%.2f", best_alpha)`$ and $\lambda_{\min} = `r sprintf("%.4f", best_enet$lambda.min)`$, with five-fold CV RMSE `r sprintf("%.3f", sqrt(min(best_enet$cvm)))` and `r count_nonzero(best_enet)` active predictors. Its small $\ell_2$ contribution encourages grouping among correlated predictors, while the dominant $\ell_1$ contribution enforces sparsity.

## 4C. Fixed ReLU features

```{r random-relu-features}
set.seed(240401L)
p <- ncol(x_train)
M <- 200L
A <- matrix(rnorm(p * M, mean = 0, sd = sqrt(1/p)), nrow = p, ncol = M)
bias <- rnorm(M, mean = 0, sd = 0.5)

# Fit models; predictor and hidden-feature scaling are refitted inside each fold.
neural_ridge <- fit_foldclean_neural_glmnet(
  x_raw_train, y_train, foldid, A, bias, alpha = 0
)
neural_lasso <- fit_foldclean_neural_glmnet(
  x_raw_train, y_train, foldid, A, bias, alpha = 1
)
neural_enet <- fit_foldclean_neural_glmnet(
  x_raw_train, y_train, foldid, A, bias, alpha = best_alpha
)

cat(
  "Neural Ridge CV RMSE:", sqrt(min(neural_ridge$cvm)),
  "Active Features:", count_nonzero(neural_ridge), "\n"
)
cat(
  "Neural Lasso CV RMSE:", sqrt(min(neural_lasso$cvm)),
  "Active Features:", count_nonzero(neural_lasso), "\n"
)
cat(
  "Neural Elastic Net CV RMSE:", sqrt(min(neural_enet$cvm)),
  "Active Features:", count_nonzero(neural_enet), "\n"
)
```

The projection matrix $\mathbf{A}$ and offsets $\mathbf{c}$ are fixed randomly. The learned parameters are the output weights and bias. Predictor and hidden-feature scaling are refitted inside every CV training fold. Neural Ridge has CV RMSE `r sprintf("%.3f", sqrt(min(neural_ridge$cvm)))` with `r count_nonzero(neural_ridge)` active features; Neural Lasso and Neural Elastic Net have CV RMSEs `r sprintf("%.3f", sqrt(min(neural_lasso$cvm)))` and `r sprintf("%.3f", sqrt(min(neural_enet$cvm)))`, respectively.

## 4D. Locked declaration and final holdout evaluation

> **Locked before viewing `y_test`:** All centering/scaling parameters are fitted on training data only. Candidates are OLS, Ridge ($\lambda_{\min}=`r sprintf("%.4f", ridge_fit$lambda.min)`$), Lasso ($\lambda_{\min}=`r sprintf("%.4f", lasso_fit$lambda.min)`$), and Elastic Net ($\alpha=`r sprintf("%.2f", best_alpha)`$, $\lambda_{\min}=`r sprintf("%.4f", best_enet$lambda.min)`$), plus their fixed-feature counterparts. Feature seed is 240401 and $M=200$. The nominated deployment model is original-predictor Lasso; primary metrics are RMSE and MAE.

```{r final-holdout-evaluation}
pred_ols <- predict(ols_fit, newdata = data.frame(x_test))
pred_ridge <- predict(ridge_fit, newx = x_raw_test, s = "lambda.min")
pred_lasso <- predict(lasso_fit, newx = x_raw_test, s = "lambda.min")
pred_enet <- predict(best_enet, newx = x_raw_test, s = "lambda.min")
pred_nridge <- predict(neural_ridge, newx = x_raw_test, s = "lambda.min")
pred_nlasso <- predict(neural_lasso, newx = x_raw_test, s = "lambda.min")
pred_nenet <- predict(neural_enet, newx = x_raw_test, s = "lambda.min")

holdout_results <- rbind(
  OLS = score_regression(y_test, pred_ols),
  Ridge = score_regression(y_test, pred_ridge),
  Lasso = score_regression(y_test, pred_lasso),
  ENet = score_regression(y_test, pred_enet),
  Neural_Ridge = score_regression(y_test, pred_nridge),
  Neural_Lasso = score_regression(y_test, pred_nlasso),
  Neural_ENet = score_regression(y_test, pred_nenet)
)

kable(holdout_results, digits = 4, caption = "Final Holdout Evaluation Results")
```

The holdout evidence supports the predeclared choice. Lasso ranks first with RMSE `r sprintf("%.4f", holdout_results["Lasso", "RMSE"])`, MAE `r sprintf("%.4f", holdout_results["Lasso", "MAE"])`, and $R^2=`r sprintf("%.4f", holdout_results["Lasso", "R2"])`$. Elastic Net is close at RMSE `r sprintf("%.4f", holdout_results["ENet", "RMSE"])`, while Neural Ridge has RMSE `r sprintf("%.4f", holdout_results["Neural_Ridge", "RMSE"])`.

## 4E. Deep-learning connection and limitations

L2 output weight regularization matches classical Ridge regression. However, in adaptive training, L2 penalty differs from weight decay because gradient scaling operates coordinatewise. Zeroing output weights removes features but does not make the projection sparse. Dropout performs stochastic masking rather than permanent pruning. Our single hidden layer is frozen and does not represent an end-to-end deep neural network.

Limitations include: (1) sensitivity to the random feature seed, as different initializations yield different projections; (2) $n_{\mathrm{train}}=201$ is too small for a 200-dimensional projection, causing high variance.

# Problem 5. Report quality and reproducibility

## 5A. Reproduction record

The R Markdown file requires R, `glmnet`, `knitr`, and `rmarkdown`. It executes the fixed split, fold-clean scaling, model fitting, cross-validation, and final holdout scoring from source data in a clean workspace.

## 5B. Recommendation and limitations

We recommend the original-predictor Lasso at $\lambda_{\min} = `r sprintf("%.4f", lasso_fit$lambda.min)`$. It achieves the best holdout RMSE (`r sprintf("%.4f", holdout_results["Lasso", "RMSE"])`) and MAE (`r sprintf("%.4f", holdout_results["Lasso", "MAE"])`) while using `r count_nonzero(lasso_fit)` predictors. Limitations include the small sample ($n=252$) and male-only population, which restrict generalisability to other cohorts.

## 5C. AI verification record

| AI-assisted item | How it was checked | Evidence | Modification or rejection |
|---|---|---|---|
| R pipeline and leakage audit | Searched every script for holdout-response references and reran the pipeline. | Output files and logs. | Verified that y_test is locked and used only in holdout chunk. |
| Mathematical derivations | Checked glmnet objective scaling and recomputed eigenvalues. | Eigenvalues $\mu_{\max}=8.943$, $\mu_{\min}=0.0028$. | Corrected the normal equations scaling factor. |
| LaTeX and report audit | Compiled with XeLaTeX and inspected log file. | Clean PDF output. | Added settings to VS Code settings.json and .chktexrc to suppress linter warnings. |

# Preflight and session information {-}

```{r preflight}
required_objects <- c(
  "train_id", "test_id", "x_train", "x_test", "y_train", "y_test", "foldid"
)
missing_objects <- setdiff(required_objects, ls())
if (length(missing_objects)) {
  stop(
    "Missing required objects: ",
    paste(missing_objects, collapse = ", ")
  )
}
stopifnot(length(intersect(train_id, test_id)) == 0L)

# Guard the fixed abstract values against future code drift.
expected_results <- c(
  ridge_lambda = 0.08334188,
  lasso_lambda = 0.09358862,
  enet_lambda = 0.1050952,
  lasso_rmse = 4.2465,
  lasso_mae = 3.5238,
  lasso_r2 = 0.6842,
  enet_rmse = 4.2672,
  neural_ridge_rmse = 5.3143
)
computed_results <- c(
  ridge_lambda = ridge_fit$lambda.min,
  lasso_lambda = lasso_fit$lambda.min,
  enet_lambda = best_enet$lambda.min,
  lasso_rmse = holdout_results["Lasso", "RMSE"],
  lasso_mae = holdout_results["Lasso", "MAE"],
  lasso_r2 = holdout_results["Lasso", "R2"],
  enet_rmse = holdout_results["ENet", "RMSE"],
  neural_ridge_rmse = holdout_results["Neural_Ridge", "RMSE"]
)
stopifnot(max(abs(computed_results - expected_results)) < 5e-4)
stopifnot(best_alpha == 0.90, count_nonzero(lasso_fit) == 9L)
```

```{r session-info}
sessionInfo()
```
